\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-07 misc/JSS article/jss-article-rnw/article.Rnw
Adding \proglang for C/C++ etc.

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures, and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference, and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \class{tibble}, \class{data.table}, \class{sf}, \class{xts}, \class{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.} \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way. \newline

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

 \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following example calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}. The latter requires an additional C routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another very neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post},\footnote{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html} vectorizes an expression to compute the $p$~value across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline

@@

\pkg{collapse} also offers higher-dimensional between and within transformations, powered by C++ code conditionally imported (and accessed directly) from \pkg{fixest}. The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.

@@

This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.\footnote{A lot may be said about benchmarking \pkg{collapse}, which would be beyond the scope of this article. Users should note, however, that its defaults (\code{na.rm = TRUE, sort = TRUE, stable.algo = TRUE, nthreads = 1}) cater to convenience rather than maximum performance. For maximum performance, set these 3 settings to \code{FALSE} and increase the number of threads. To also provide a minimalistic guide for \proglang{R} users seeking to understand the relative performance of \pkg{collapse} and \pkg{data.table}, reflecting current (spring 2024) developments: \pkg{collapse} has highly efficient algorithms for grouping and computing statistics, but presently does not provide sub-column level parallel grouping architecture. Simple statistics like \code{fmean()} are parallelized across columns and perform grouped computations in a single pass. More complex ones \code{fmedian(), fmode()} have group-level parallelism. \pkg{data.table}, on the other hand, has sub-column parallel grouping and also group-level parallel implementations for simple statistics such as \code{mean()}, but no parallelism for complex statistics such as \code{median()}. \pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}. Thus, if your data is moderately sized ($\leq$100mio. obs.), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice. On the other hand, if your data is really long ($>$100mio. obs.), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice. My recommendation: use both, just need to call \code{library(fastverse)}. Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language. While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since C, and particularly R's C API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers. At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database. \pkg{collapse} seamlessly integrates with \proglang{R}'s data structures; \pkg{polars}, at present, has nothing to do with them (and is therefore also not part of this benchmark).}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures, and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference, and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \class{tibble}, \class{data.table}, \class{sf}, \class{xts}, \class{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.} \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function. However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way. \newline

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

 \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following example calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}. The latter requires an additional \proglang{C} routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in \proglang{C} to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another very neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post},\footnote{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html} vectorizes an expression to compute the $p$~value across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline

@@

\pkg{collapse} also offers higher-dimensional between and within transformations, powered by \proglang{C++} code conditionally imported (and accessed directly) from \pkg{fixest}. The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.

@@

This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.\footnote{A lot may be said about benchmarking \pkg{collapse}, which would be beyond the scope of this article. Users should note, however, that its defaults (\code{na.rm = TRUE, sort = TRUE, stable.algo = TRUE, nthreads = 1}) cater to convenience rather than maximum performance. For maximum performance, set these 3 settings to \code{FALSE} and increase the number of threads. To also provide a minimalistic guide for \proglang{R} users seeking to understand the relative performance of \pkg{collapse} and \pkg{data.table}, reflecting current (spring 2024) developments: \pkg{collapse} has highly efficient algorithms for grouping and computing statistics, but presently does not provide sub-column level parallel grouping architecture. Simple statistics like \code{fmean()} are parallelized across columns and perform grouped computations in a single pass. More complex ones \code{fmedian(), fmode()} have group-level parallelism. \pkg{data.table}, on the other hand, has sub-column parallel grouping and also group-level parallel implementations for simple statistics such as \code{mean()}, but no parallelism for complex statistics such as \code{median()}. \pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}. Thus, if your data is moderately sized ($\leq$100mio. obs.), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice. On the other hand, if your data is really long ($>$100mio. obs.), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice. My recommendation: use both, just need to call \code{library(fastverse)}. Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language. While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since \proglang{C}, and particularly \proglang{R}'s \proglang{C} API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers. At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database. \pkg{collapse} seamlessly integrates with \proglang{R}'s data structures; \pkg{polars}, at present, has nothing to do with them (and is therefore also not part of this benchmark).}
